Expanding the scope of experimental archaeology using the Perception-Process-Product analytical framework

### 2022-12-05 manuscript/manuscript.Rmd
updated the outline of this article.

This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)). Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016]. In the past decades, actualistical experiments becomes more common [@liu2022].

@@

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is innovative yet limited by the curse of expertise [@hinds1999]. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014].

@@

>>>>

This paper presents the Perception-Process-Product analytical framework to expand the scope of experimental archaeology (**Figure** \@ref(fig:concept)). minimal engineering using raw materials available in the past to demonstrate it is possible to do something.

This part talks about the goal and toolbox of PPP framework, which is understanding the multi-level understanding of variation. the first two p captures different level of variation

Traditionally, experimental archaeology focuses on generating knowledge regarding the causal mechanism at behavioral level to explain the variation of material culture [@eren2016]. In the past decades, actualistical experiments becomes more common [@liu2022].

@@

Ethological approaches has been first systematically developed and applied in the archaeological research by Haidle [@haidle2009; @haidle2010; @lombard2012], known as cognigram, essentially representing an abstracting process of a series of behavioral sequences achieving a similar goal. This approach is a power and elegant yet limited by the curse of expertise [@hinds1999]. Like chaine operatoire, it cannot handles variation very well. To some extent it describes the minimal steps to achieve a goal from the perspective of reverse engineering and assume clear causal thinking between each steps. Novices has a different sets of perception on the causal structure of how certain behaviors will modify the raw materials, leading to over-imitation. Here we used the ethogram, or the action grammar, developed by [@stout2021] as an example. Other coding scheme also exist such as [@mahaney2014].

# The curse of knowledge

Variation: why novice is important?

@@

Variation: why experts or collaborative knowledge from different regions matters?
